---
title: "R for Bio Data Science"
author: "Oriade Latifah Simpson"
date: "Spring 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Group Number: 20**

**Instructions:**

Task 2.6: Your GROUP ASSIGNMENT is any written document (might be even a simple “.txt” file) containing the link to your repository and name of the file used in this exercise. We can inspect it by using blame feature, which shows who committed , and what was commited.

***

### Scripting in a reproducible and collaborative framework using Github Via Rstudio

***

### Introduction

During these exercises we are going to learn how to create a Github repository and how to use it to synchronise work between all team members. 
It is split into a few parts:

1. Setting up a repository
2. Basic Committing and Synchronising. 
3. Solving Conflicts
4. Checking out and basic branching. 


The first three are the most essential for working with Git and Github. The last explains how to use git features to speed up the development process. 
Use the learning materials to explain how to perform Git commands in RStudio. 

Rules:

* Use the RStudio Interface Only. 
* Communicate clearly with your team and also remember to listen them. 
* If you make a mistake, use the emergency options. 


Emergency Options:

* If you made mistake, use the revert command.
* If you cannot pull, copy the files somewhere else, reset then try to pull. If this doesn't work then stash the index.
* If you cannot push, make sure to pull first. 
* If it does not work, start a new R project and pull all of the changes. 

***

### Setting up a repository

Do you have a github account? Yes. 

1. The Github website is : <https://github.com/>

2. Select rforbiodatascience22 as the owner , create a new repository called  group-20-git-tutorial where 20 is your group number. 
   Select the public option, add a README.md file and click CREATE. 
   
3. Go to repository Settings (bar above the files), select Access -> Collaborators and hit Add people > add names of your team mates. 

4. See if the others accept the invitation. 
(yes)

***

### Clone a Repository on RStudio Cloud 

1. Go to the RStudio Cloud

2. Select our workspace (F2022 R For Bio Data Science)

3. Upper right corner select New Project > New Project from Git Repository. 
Paste the URL of your repository

https://github.com/REPO-OWNER-USERNAME/REPO-USERNAME e.g.

**https://github.com/rforbiodatascience22/group-20-git-tutorial**
                             
Congrats! You all have a copy of a repository in your project!

This action of setting up a copy of a repository is called cloning.

***

### Set up a username

This should be done every time you set up a new project in RStudio cloud.

Set up your GitHub name and email for the project. 

You do it by typing in the console:

```{r}
# install the usethis package first. 
#install.packages("usethis")
library(usethis)
```

```{r eval=FALSE}
usethis::use_git_config(user.name = "username", user.email = "email_address")
```

***

### Set up a PAT

Either configure SSH connections or use Personal Access Token (PAT).

PAT is like a password, for interacting with GitHub via RStudio Cloud. 

Generate it via the GitHub page.

Type in the console :
```{r eval=FALSE}
usethis::create_github_token()
```

1. GitHub website > type your password > add a description – set ‘RStudio Cloud’ > hit Generate token.

Save the password securely.

When trying to pull or push from remote repository use the PAT. 

Manual setting up the PAT: 

1. Type in the console:

```{r eval=FALSE}
gitcreds::gitcreds_set(). 
```

2. Paste the PAT when it says Enter password or token. 

Managing Github credentials information : <https://usethis.r-lib.org/articles/git-credentials.html> 

***

### Commit the first files

1. Open Git in your RStudio Project stage two files called .gitignore and .Rproj

2. Commit the files and add a message. 

3. Push the changes. 

(2 Files Changed 17 insertions.)

.gitignore is a special file, which states which other files in the project should not be committed. 
.Rproj is a project configuration file. 

***

### Synchoronise the repository

Others do as above. 

***

### Basic Commiting and Synchronising

***

### Create and Commit a File

1. Each team member: Create files in RStudio. Fill the file/s with any content. Save it with a unique name among your team. Pull from the remote repository, create, stage and commit the file/s, push the changes.

2. Other team member: Pull from the remote repository,
then create normally, stage and commit your own files and push the changes.

3. All: Browse through the commit history, compare with remote repository history 
(can be found on GitHub webpage).

Diff > History. 

***

### Concurrent Modifications

This is the simplest workflow of using git: pull, commit, push.
As long as you do not modify the same file you can use it to synchronize your work.


**ONE AT A TIME:**

1. One person: Pull, modify one or two of the files 
+ (e.g. add a line, remove a line, modify a line, stage, commit and **push**.


2. Other person:  Pull, rename one file, commit and **push**.
The file shows as deleted. An error will pop up asking to stash changes .

Type: 

```{bash eval=FALSE, include=FALSE}
git stash
```

This deletes the changes. 

3. Other person:  Pull, delete one file, push.

Note: When RStudio Cloud goes to sleep, refresh the page.

***

### Git Ignore

.gitignore is a very special file because it states which files in the repository should not be committed. 

Exclude files that include passwords or pdf results. If you have a reproducible workflow, commit the “.Rmd” file, but ignore the “.pdf”, as anyone can rerun it and get the same result.

<https://www.atlassian.com/git/tutorials/saving-changes/gitignore> is a tutorial for the “.gitignore” file.

1. One person: Create a file named “clever_file.txt”. 

Add a line “clever_file.txt” to your “.gitignore” file. 
Commit & push it. 

Notice there is no option to stage the “clever_file.txt”. 
Note: Other members cannot see this file because it is in the .gitignore. 


Others: **Pull**. Create your own “hidden_file.txt”.
Add a line “hidden_file.txt” to your “.gitignore” file. 
Commit & push it. 


One Person: Create a “document” directory. Create an “Waterfalls.Rmd” document with any content in the “document” directory, but do not knit it. Commit it.

Knit the document.
Add a line “Waterfalls.pdf” to your “.gitignore” file. 
Commit changes in the “.gitignore”. Push it.

Others: Pull.
Note: Other members cannot see these files because they are in the .gitignore. 

**Note: There is a long delay on github with getting the changes to be shown online.**

***

### Guardians of the directory

Another special file is .gitkeep.

Create a directory in your local repository. 
You cannot stage it, as you could do with regular files.

If you really want to keep a folder without its content, but you want to keep a folder for synchronization.
(e.g. big files which you do not want to push to a remote repository) 


One person: Create a folder called “abc”. Inside create an empty file called “.gitkeep”. Commit and push it.

All team members: Pull.
Note: All team members will see there is nothing in the abc folder. 

One person: In the directory “abc” create a file called “XYZ.csv” without content inside.
a) Don’t commit the file.
b) Add it to “.gitignore” regarding the path (it should be “abc/XYZ.csv”). 
c) Commit the “.gitignore” and push.

Note: If you make a mistake of committing this file, everyone will see it when you push. Instead, cache the file to stop it from being tracked. 

```{bash eval=FALSE}
git rm --cached abc/XYZ.csv
```


All team members: Pull.

Right now you could write code that, e.g. downloads a file called “XYZ.csv” to a folder “abc” and you are sure that the folder exists and the file itself won’t get committed.

Note: The XYZ file is not staged.

***

### README.md Files

Very common file for a repository is README.md. This file is usually a showcase of your repo and GitHub (as you might have noticed it) displays it on its website. 

One team member: Modify it so that it informs others why the repository exists and who contributes to it.

Source:

How to use Git Ignore : <https://www.pluralsight.com/guides/how-to-use-gitignore-file>

***

### Modify the same file

This task is meant to show you that in fact you can modify one file at the same time.

1. One team member: Pull. Create a new file consisting of 10 empty lines. Commit and push.

Note: This file is called: **Ten_Empty_Lines.R**

2. Other: Pull. Modify one line of the file. Make sure that each of you modifies a distinct line.

3. Other: Make a change, then commit the changes without pulling, then compare histories.

Note: This will give an error message of:
Updates were rejected because the remote contains work that you do
hint: not have locally. 


Other: Pull and push. Notice the message you got when pulling (except for the first one of you).

Other: Pull to make sure to have the same state of the repository.

Note: The final file will be changed properly. 

***

### The Actual Task 

Create a written document containing the link to your repository and name of the file used in this exercise. 

The teachers can inspect it by using blame feature, which shows who commited which files, and what was commited

The link to the repository is: <https://github.com/rforbiodatascience22/group-20-git-tutorial>

The name of the file used in this exercise is: **Ten_Empty_Lines.R**

***


