---
title: "R for Bio Data Science"
author: "Oriade Latifah Simpson"
date: "Spring 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Group Number: 20**

***

**Instructions**

Let's go with PDF again.

Remember, this is your chance to practise your stakeholder communication, so add a title, group number and names, some headers, a bit of text, etc.

***

Information about the data is found via this link here : <https://hbiostat.org/data/repo/Cdiabetes.html> 


***

**Diabetes Dataset**

These data are courtesy of Dr John Schorling who worked at the Department of Medicine, University of Virginia School of Medicine.

The data consist of 19 variables on 403 subjects from 1046 subjects who were interviewed in a study to understand the prevalence of obesity, diabetes, and other cardiovascular risk factors affecting African Americans in central Virginia, USA. 

According to Dr John Hong, Diabetes Mellitus Type II (adult onset diabetes) is associated most strongly with obesity.

The waist/hip ratio may be a predictor in diabetes and heart disease. 
Diabetes Mellitus Type II is associated with hypertension. In this study, 403 subjects were screened for diabetes. 
Glycosolated haemoglobin levels that were measured to be higher than 7.0 were recorded as being a positive diagnosis of diabetes. 

For more information about this study see:

* Willems JP, Saunders JT, DE Hunt, JB Schorling: Prevalence of coronary heart disease risk factors among rural blacks: A community-based study. Southern Medical Journal 90:814-820; 1997 


* Schorling JB, Roach J, Siegel M, Baturka N, Hunt DE, Guterbock TM, Stewart HL: A trial of church-based smoking cessation interventions for rural African Americans. Preventive Medicine 26:92-101; 1997.

***

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(ggthemes)
library(patchwork)
```

***

**Q18: Which correlate better with BMI, WHR or BFP?**

**Is there a certain plot-type, which can visualise if two variables are correlated?**

**Can you perhaps use an R-function to compute the “correlation coefficient”?.**

**Do not use e.g. ggpubr, use only tidyverse and base)**

```{r message=FALSE, warning=FALSE}
diabetes_data <- read_csv(file = "data/diabetes.csv", na = "NA")
```


```{r}
diabetes_data_t1 <- diabetes_data %>% 
  mutate(height_cm = round((height * 2.54), digits = 2),
         weight_kg = round((weight / 2.20), digits = 2),
         waist_cm = waist * 2.54,
         hip_cm = hip * 2.54,
         height_metres = height_cm /100,
         gender = factor(gender),
         location = factor(location),
         bmi = round(weight_kg / (height_metres ^ 2), digits = 1),
         gender_class = case_when( gender == "female" ~ 0,
                                    gender == "male" ~ 1),
         body_fat_percentage = ((1.39 * bmi) + (0.16* age) - (10.34 * gender_class) - 9),
         waist_to_hip_ratio = round((waist_cm/ hip_cm), digits = 2))

```

***

**BMI and waist to hip ratio**

```{r}
# drop the na values. 
drop_na(diabetes_data_t1) %$%
  cor(bmi, waist_to_hip_ratio) 
```

The correlation for bmi and waist to hip ratio is -0.08


```{r message=FALSE, warning=FALSE}
#Plot the relationship between bmi and waist to hip ratio**

pl1 <- diabetes_data_t1 %>%
  
  ggplot(aes(x = bmi, y = waist_to_hip_ratio, fill = bmi)) +
  
  # change the size/transparency of the points
  geom_point(size = 3, colour = "black", alpha = 0.5, pch = 21)+
  
  # fill with a continuous colour. 
  scale_fill_continuous()+
  
  labs(subtitle = "bmi and waist to hip ratio correlation coefficient: -0.08",
       x = "body mass index",
       y = "waist to hip ratio")+
  
  geom_smooth(method = "lm", colour = "skyblue2") +
  
  theme_clean()
 

``` 


***

**BMI and Body Fat Percentage**


```{r}
drop_na(diabetes_data_t1) %$% 
  cor(bmi, body_fat_percentage)
```

The correlation for bmi and body fat percentage is 0.880

```{r message=FALSE, warning=FALSE}
#Plot the relationship between bmi and body fat percentage

pl2 <- diabetes_data_t1 %>%
  
  ggplot(aes(x = bmi, y = body_fat_percentage, fill = bmi)) +
  
  # change the size/transparency of the points
  geom_point(size = 3, colour = "black", alpha = 0.4, pch = 21)+
  
  # fill with a continuous colour. 
  scale_fill_continuous()+
  
  labs( subtitle = "bmi and body fat percentage correlation coefficient: 0.88",
        x = "body mass index",
        y = "body fat percentage")+
  
  geom_smooth(method = "lm", colour = "darkgreen", linetype = "solid") +

  theme_clean()
``` 

***

Use the patchwork library to put the plots together.

```{r message=FALSE, warning=FALSE}
pl1 / (pl2)
```


The general trend of body mass index and waist to hip ratio is negative with a correlation coefficient of -0.08. 
which indicates a non-linear relationship between the two variables.

The general trend of body mass index and body fat percentage is positive and the correlation coefficient 0.88
indicates the degree of linear relationship between the two variables. 

***


