---
title: "R for Bio Data Science"
author: "Oriade Latifah Simpson"
date: "Spring 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Group Number : 20**

***

**Instructions**

For this assignment you are to apply what you have learned in the two data visualisation labs.

The task is to create a really nice plot using one of two datasets, the cancer_data or the SPE_ENV

This time, submit one PDF file per group.

Submit only this assignment, not the entire exercise

Make it nice, add a title, your group number and names and a bit of text to along with your data visualisation

Deadline is Thursday 23:59 as usual

***

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggridges)
```


```{r}
SPE_ENV = read_tsv(file = "data/SPE_ENV.tsv")
SPE_ENV
```


```{r message=FALSE, warning=FALSE}
# pick a few taxa. 
new_taxa <-  c("Flavobacteria", "Bacteroidia", "Acidobacteria_Gp18", "Methaobacteria")

SPE_ENV2 <- filter(SPE_ENV, Taxa %in% new_taxa)
#SPE_ENV2

SPE_ENV2 %>% 
  ggplot(mapping = aes(x = OTU_Count, y = Taxa, fill = Taxa)) +
  
  # draw the quantiles in the distribution. 
  stat_density_ridges(quantile_lines = TRUE, alpha = 0.5,
                      
  # Add a rug underneath. 
  jittered_points = TRUE,
  position = position_points_jitter(width = 0.05, height = 0),
  point_shape = '|', point_size = 3, point_alpha = 1)+

  # change the limits of the x axis
  scale_x_continuous(limits = c(-250, 1000), expand = c(0, 0)) +

  facet_wrap(~site)+
  
  theme_gray()+

  labs(
    x = "Operational Taxonomic Unit Count",
    y = "Taxa Selected",
    title = "Layered Densitogram",
    subtitle = "Distribution of Operational Taxonomic Units of selected bacteria",
    caption = "Data from doi.org/10.1111/1751-7915.12334") +
  
  scale_fill_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))+
  
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 60, hjust=1, size = 7),
    plot.title=element_text(face="bold", size=20, colour = "darkblue"))


```

***

* According to the research paper below, the operational taxonomic unit represents species richness.

* The density plots above show that there is generally a higher variance in the "species richness" of Flavobacteria, Bacteroidia and Acidobacteria_GP18 in Vietnam, than there is in Tanzania.

* These bacteria were sampled from 30 pit latrines in two different geographical regions; Tanzania and Vietnam. 

* Paper: **Assessment of the influence of intrinsic environmental and geographical factors on the bacterial ecology of pit latrines.**

The link to the paper can be found here : <https://sfamjournals.onlinelibrary.wiley.com/doi/full/10.1111/1751-7915.12334>

***


