---
title: "R for Bio Data Science"
author: "Oriade Latifah Simpson"
date: "Spring 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Group Number 20**

***

**Instructions**

Here is the assignment for lab 5. Make sure to read and implement any comments you may have gotten as feedback on the lab 4 hand in.

Due Date 3 March 2022 @ 11.59pm

***

A recent study of T-cell receptors and SARS-CoV-2
<https://www.researchsquare.com/article/rs-51964/v1>

* The data is available from this link: 
<https://clients.adaptivebiotech.com/pub/covid-2020>

* There are also other data available for open analyses on this website: 
<https://clients.adaptivebiotech.com/immuneaccess/browse?>

* Release 002.1 - over 160,000 high-confidence SARS-CoV-2-specific TCRs

* The downloaded zip file is ImmuneCODE-MIRA-Release002.1.zip

* The two main files are called peptide-detail-ci.csv and subject-metadata.csv are saved in the data folder. 

***

 
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r message=FALSE, warning=FALSE}
peptide_data <- read_csv("data/ImmuneCODE-MIRA-Release002.1/peptide-detail-ci.csv",
                         na = c("", "NA", "N/A"))

meta_data <- read_csv("data/ImmuneCODE-MIRA-Release002.1/subject-metadata.csv",
                      na = c("", "NA", "N/A"))
```

***

**Q15: Per gene and in terms of percent, what are the 3 top alleles?**
**Recreate this plot to answer the question**

```{r}
peptide_meta_data <- full_join(peptide_data, meta_data)

# Remove columns that start with D,R,Q or P. 
peptide_meta_data <- peptide_meta_data %>% 
  select(-matches("D[RQP]"))

#peptide_meta_data
```

```{r}
allele_data <- peptide_meta_data %>% 
  select(Subject, matches("H[LA]")) %>% 
  distinct(Subject, .keep_all = TRUE)

#allele_data
```

**Use pivot longer to convert the data into a long format.**

```{r}
allele_data_long <- allele_data %>%
  pivot_longer(
    cols = starts_with("HLA"),
    names_to = "Gene",
    values_to = "Allele",
    #values_drop_na = TRUE
  )

#allele_data_long
```

**Q12: How many unique alleles are there?**


* There are 155 HLA Alleles however, A*01:01 and A*01:01:01 are not in the same category. 

* The type comes after the first *, so the type is 02, 23 44 , 04, 05, 11, 07 or 15 etc. 

* The letter A, B or C stands for the gene. Gene A, Gene B, Gene C.

* Field 2 comes after the semicolon and stands for the specific HLA protein, which is 13, or 01, or 02.

* Afterwards, there may be a substitution or difference in a non-coding region and then a letter. 

* The letter denotes change in the expression e.g N represents an allele that is not expressed. 


```{r}
allele_data_long %>% 
  group_by(Allele) %>% 
  summarise(n = n())
```

```{r}
allele_data_long <- allele_data_long %>% 
  # cut off the numbers from the Gene. 
  mutate(Gene = str_sub(Gene, 1, 5),
         # Collect the first 7 characters of the allele. 
         Allele_F_1_2 = str_sub(Allele, 1,7),
         # turn the gene into a factor and remove the NAs
         Gene = as.factor(Gene)) %>% 
         drop_na()

#allele_data_long
```

**Q13: How many unique alleles are there now?**

There are now 110 unique alleles. 

```{r}
allele_data_long %>% 
  group_by(Allele_F_1_2) %>% 
  summarise(n = n())
```

**Q14: What are the top 3 alleles in terms of counts in the data?**

The top three alleles are :

* A*02:01 with 75 with a high 10.9 % 

* C*07:02 with 43 with 6.28 %

* B*07:02 with a close 33 coming in at 4.82 %

```{r}
allele_data_long %>% 
  count(Allele_F_1_2)%>%
  arrange(desc(n)) %>% 
  mutate(percent = n/sum(n) * 100) %>% 
  head(n = 3)
```



**Q15: Per gene and in terms of percent, what are the 3 top alleles?**
**Recreate this plot to answer the question GROUP ASSIGNMENT**

```{r}
plot_end_data <- allele_data_long %>% 
  group_by(Gene)%>%
  count(Allele_F_1_2)%>%
  arrange(desc(n)) %>%
  mutate(percent = n/sum(n) * 100) %>% 
  arrange(desc(percent))


#plot_end_data
```

```{r}
# Factor the genes based on the group_by
# Make the y axes similar using free_y and generate 3 columns using facet wrap. 

plot_end_data %>% 
  ggplot(mapping = aes(x = percent, y = fct_reorder(Allele_F_1_2, percent))) +
  geom_col()+
  
  # rather than separating the data, just use facet wrap. 
  
  facet_wrap(vars(Gene), scales = "free_y", ncol = 3)+
  labs(x = "Observed Allele Frequency (Within each gene) [%]", 
       y = "Allele")+
  theme_minimal(base_family = "Times", base_size = 9)

```

***

* Source: A large scale database of T-cell receptor beta sequences and biding associations from natural and synthetic exposure to SARS-CoV-2 <https://www.researchsquare.com/article/rs-51964/v1> 


* Nomenclature for Factors of the HLA system:<http://hla.alleles.org/nomenclature/naming.html>

***


